Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment

### 2023-11-20 revised_manuscript_231115.Rmd
uploading replication datasets

---
title: "Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^,  Maike Hamann^6,7^, Emilie Lindkvist^2^, Jiangxiao Qiu^8^, 
    | \small Caroline Schill^2,9^, Alon Shepon^10^, Anne-Sophie Crépin^9^, and Carl Folke^2,9^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^Centre for Geography and Environmental Science, University of Exeter
    | \footnotesize ^8^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^9^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^10^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: paper/references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

@@

- Historical data reveals that no country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact.
- A sustainability trilemma seems to exist where progress on either social goal comes at the expense of the environment.
- An inequality trap seems to be particularly acute in Latin America.
- Country trajectories in the trilemma space suggest that future social progress can be achieved without compromising the biosphere.

@@

We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support. This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.

@@

# Global tradeoffs between prosperity, inequality, and the environment

@@

  A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. But what if there is a "sustainability trilemma" that inhibits the simultaneous achievement of these three objectives? We analyzed synergies and tradeoffs between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017. Based on these widely-used measures of development, we found that no country overcame the hypothesized trilemma at the intersection of prosperity, equality, and environment. The trajectories of countries in the three-dimensional space formed by these key indicators revealed typologies and common development barriers. Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact. Some countries were trapped in a space of high inequality, while the instances of prosperous countries lowering inequality coincided with increasing environmental impact. The existence of clusters and widely divergent trajectories of certain countries within the trilemma space point to the need for explanatory frameworks that incorporate institutional, historical, and geographical drivers, as well as social-ecological feedbacks that strengthen their effects. Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.

@@

Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment - what is sometimes referred to as the "triple bottom line". More specifically, a key development goal within this tripartite framework is to achieve high material living standards (prosperity) that are also well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Programmatically, the United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment  [@Sachs:2015]. And in its report charting the path towards a sustainable recovery in the aftermath of COVID-19, the World Bank also identified the need to achieve "shared prosperity" by reducing inequality and accounting for the climate crisis and other environmental challenges [@WorldBank:2022]. Although each objective is desirable in its own right, whether prosperity, equality, and environmental integrity can be simultaneously achieved remains an empirically unanswered question. In other words, what if the achievement of the sustainability triple-bottom-line is impeded by the existence of a sustainability "trilemma"?

In the literature on international development, there have been many hypothesized "trilemmas" involving separately desirable but mutually incompatible objectives. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination [@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has been proposed as an unattainable triad [@Coyle:2011]. Similarly, structural tradeoffs between prosperity, equality, and the environment  have been investigated in different locations using various indicators. A statistical analysis found that economic inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries [@Grunewald2017]. In contrast, an earlier econometric study concluded that mitigating carbon emissions trades-off with both economic growth and social equity [@Ravallion_2000]. And focusing on three case studies in the United States and Southeast Asia, Halpern et al. (2013) found that achieving biodiversity objectives is often facilitated by the inequitable distribution of resources, complicating the achievement of "triple-bottom-line solutions" in conservation planning [@Halpern_2013]. Using an agent-based model of collective decision-making that assumes simple majority rule and takes for granted prosperity-environment conflicts, Kempf and Rossignol (2007) argued that widening inequality undermines public willingness to support environmental policies [@KEMPF_2007].

Despite earlier studies on potential sustainability trilemmas, few analyses have taken a dynamic global perspective. Many of the most critical development challenges in the Anthropocene, when human activities have become the primary drivers of planetary change, are transboundary and global in scale. Identifying the distribution of these challenges across countries is key to arriving at principles for collective action. At the same time, it is important to track how different countries have progressed or regressed with respect to key objectives over meaningful time-frames. In this way, country-specific as well as shared international obstacles and opportunities can be identified, helping to inform the design and implementation of policies.

In this study, we ask whether there is evidence – and rationale – that the three sustainability goals of prosperity, equality, and environmental integrity can be jointly achieved by any given country using an international, time-series dataset of representative development indicators. In other words, this study attempts to answer the questions: To what extent does such a sustainability trilemma exist and are there inherent tradeoffs between its constituent goals? If such a sustainability trilemma does exist, where does it occur and why does it occur there? By providing answers to these questions, we hope to contribute to the increasingly urgent discourse on how to achieve sustainable and inclusive prosperity in an age of unprecedented planetary change.

## Background

For most of human history, material want, not sufficiency, was the norm for the majority of people, regardless of country [@Maddison:2007; @Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018]. Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an unprecedented increase of global poverty and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for certain countries in what was once called the "Third World" have converged with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].

The widespread feelings of injustice that highly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009]. Over the past decade, inequality has once again become an increasingly urgent issue for policymakers and the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020]. These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks [@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017]. Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened economic inequalities within many countries, as well as widened the development trajectories between high- and low-income economies [@WorldBank:2022].

Environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic]. Interdisciplinary research confirms that human well-being and economic development depend on the resilience of nature. There is also a scientific consensus that human activities have a wide and deleterious impact on the biosphere, which constitutes the habitative upper bound and common denominator for our species (notwithstanding fanciful visions of extraterrestrial expansion)  [@Díaz:2019; @folke2021our]. Anthropogenic drivers have become primarily responsible for global biodiversity loss, land degradation, and climate change, among other environmental impacts [@Díaz:2019]. Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023]. Global development in the 21st Century therefore urgently requires more sensible stewardship of the biosphere, from ecosystems to the climate [@Millenniu:2005; @IPCC:2021].

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time. We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

@@

Our analysis of the three representative indicators and their respective thresholds found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.

While these globally-aggregated patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies. To further investigate individual trajectories, we calculated the displacement within the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane. These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
\caption{\textbf{Countries' positions and trajectories within the trilemma space} 59 country trajectories from 1995 to 2017 are depicted. Colours and contours represent areas of similar values and enclose attractors within peaks. The insets of each graph show the direction of movement for countries during the study period.}
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). Building on insights from feedback dynamics described in other fields of development studies and sustainability science, multiple stable states within the trilemma space are plausible if feedbacks between development drivers are present. Statistically, this translates into multi-modal distributions and so-called "attractors". The prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three trilemma indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A). First, there was a group of countries, mainly from Latin America, that were "trapped" in a region of high inequality. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively high prosperity and ecological footprint; their movements in the trilemma space were relatively less erratic, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).

If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma indicators, we are able to include more countries (N=140). This static analysis yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the dynamic analysis. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact. The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "developing world" or "Global South". Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.

Decreasing inequality and decreasing environmental impact tended to pull countries in opposite directions (Fig \ref{fig:pca}), providing strong evidence for the difficulty of achieving both objectives at the same time. In Fig \ref{fig:pca}A the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences. While environmental footprint (EF) and gross national income (GNI) point in similar directions, meaning countries tend to increase both at the same time, our inequality proxy (Gini) points in the opposite direction, confirming our previous speculation: countries that reduce inequality do so by increasing their environmental footprint and prosperity. In Fig \ref{fig:pca}B the ordination is performed in a larger but static sample (140 countries), and we find a similar pattern although not completely in opposite directions. In fact, in our trajectories analysis we found that  countries that reduced inequality subsequently increased their environmental impact (Fig \ref{fig:cube_ensamble}A). Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B). However, some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact - albeit at lower levels of prosperity (Fig \ref{fig:cube_ensamble}B, C). Overall, our findings are in line with the conventional wisdom that higher prosperity usually comes at environmental expense. The low frequency with which countries were able to meet even two thresholds - none in the case of the prosperity-environment relationship - is notable. National trajectories through the three-dimensional trilemma space also revealed that some countries struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the more general sustainability trilemma.

```{=tex}
\begin{figure*}[htp]
\centering
\includegraphics[width = 6.5in, height = 5in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. Three clusters are identified when looking at the historical trajectory of countries (N=59) from 1995-2017 (A). Two clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other. Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

@@

In this study, we conducted a retrospective, dynamic analysis, and it is important to note that past trajectories do not necessarily determine future pathways. The potential mechanisms of tradeoffs and synergies between prosperity, inequality, and environmental impact constitute an important research frontier in sustainability science and development studies. In the often disparate literatures of economic, social, and environmental science, no theory has so far been advanced to account for the country patterns and typologies identified here. Certain mechanisms have been hypothesized for pairwise relationships among the development indicators used in this study. However, only by considering all three dimensions together, and the different ways in which countries have evolved with respect to them, can we derive a more comprehensive picture of the sustainability challenges for international development. Before positing trilateral explanations for the existence and distribution of the sustainability trilemma, we will first review several of the most influential dilemma hypotheses and discuss why they may be at odds with our findings.

Tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. The evidence for this hypothesis has been contested [@Chancel:2020]. Sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or even concerted government actions [@Pikett:2014]. Our analysis of per-capita income and income distribution found no evidence for a KC-like relationship from 1995 to 2017, which was a period of rapid economic growth in many parts of the world (see SI).

>>>>

---
title: "Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,4§^, Kevin Berry^5^, Tomas Chaigneau^6^,  Maike Hamann^7,8^, Emilie Lindkvist^2^, Jiangxiao Qiu^9^, 
    | \small Caroline Schill^2,4,10^, Alon Shepon^11^, Anne-Sophie Crépin^10^, and Carl Folke^2,4,10^
    | \small
    | \footnotesize ^1^The Natural Capital Project, Doerr School of Sustainability, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^ The Anthropocene Laboratory, The Royal Swedish Academy of Sciences
    | \footnotesize ^5^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^6^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^7^Centre for Sustainability Transitions, Stellenbosch University
    | \footnotesize ^8^Centre for Geography and Environmental Science, University of Exeter
    | \footnotesize ^9^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^10^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^11^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: paper/references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

@@

- No country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact between 1995-2017.
- A sustainability “trilemma” as opposed a “triple bottom line” appears to exist for international development.
- There is an “inequality trap” that appears to be particularly acute in Latin America.
- Country trajectories suggest that reducing inequality can be achieved without compromising the biosphere.
- The three dimensions need be considered together to identify international development challenges.

@@

We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support. This work was also supported by funding from Formas (grant number 2018-02318 and grant number 2020-00454), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.

@@

# Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment

@@

A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. This has often been called the “triple bottom line”. But instead, what if there is a " trilemma" that inhibits the simultaneous achievement of these three goals? We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017. Based on these widely-used measures of development, we found that no country achieved the triple bottom line. Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact. A cluster of countries, primarily from Latin America, appeared trapped in a regime of high inequality, while the instances of prosperous countries lowering inequality only coincided with increasing environmental impact. The existence of country clusters and widely-divergent national trajectories show that common pair-wise explanations such as Kuznets Curves do not fully capture important development dynamics, especially in a period of global environmental change. Instead, frameworks need to consider all three dimensions together, and especially the relationship between inequality and the environment. Additionally, the past development trajectories of countries, how they fit into the structure of the world economy, and the role of social-ecological feedbacks in strengthening clustering effects such as “inequality traps”. Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.

@@

Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment - what is sometimes referred to as the "triple bottom line". More specifically, an important development objective within this framework is to achieve high material living standards (prosperity) that are also well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Programmatically, the United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment [@Sachs:2015]. And in its report charting the path towards a sustainable recovery in the aftermath of COVID-19, the World Bank also identified the need to achieve "shared prosperity" by reducing inequality and addressing environmental crises [@WorldBank:2022]. However, whether the three goals can be simultaneously achieved remains an open question – what if there is a "sustainability  trilemma" that prevents countries from reaching this development objective?

Although global prosperity has increased in recent generations, progress has been uneven. Even today, a significant share of the world’s population lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold, hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018]. And while average incomes for certain developing countries have been converging with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened across the world [@Bourguignon:2007; @Milanovic:2016]. This has made  inequality an urgent issue for policymakers and the general public due to  anxieties about social and political instability [@Chancel:2020; @UN2020]. Environmental integrity forms the third, and perhaps most fundamental, dimension of the sustainability [@arrow1995economic]. There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Díaz:2019; @folke2021our]. Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].

Identifying global patterns and trends in the relationships between prosperity, equality, and environmental impact can provide insights into the development challenges facing individual  countries as well as shared obstacles and opportunities internationally. Using a time-series dataset of representative development indicators, this study asks whether there is evidence that prosperity, equality, and environmental integrity has been jointly achieved by any country, and then explores potential mechanisms of trade-offs and synergies between the three goals. In other words, this study attempts to answer the questions: To what extent does a "sustainability trilemma" exist, and if does exist, where does it occur and why does it occur there? By providing answers to these questions, we hope to contribute to the increasingly urgent discourse on how to achieve sustainable and inclusive prosperity in an age of unprecedented planetary change.

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the three-dimensional space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time. We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

@@

Our analysis found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.

While these global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies. To further investigate country trajectories, we calculated the displacement within the three-dimensional space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane. These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 7in, height = 5in]{trilema_ensamble2}
\caption{\textbf{Countries' positions and trajectories within the trilemma space} 59 country trajectories from 1995 to 2017 are depicted. Colours and contours represent areas of similar values and enclose regimes (attractors). The bottom panel shows the direction of movement for countries during the study period for each facet of the trilemma cube. Angles and distances are calculated in the three-dimensional space, but visualised in two dimensions for readability. The arrows shows the direction and magnitude of movement if all countries would have started at the same point of origin (to enable comparison). Arrow colours on the bottom panel correspond to groups found on the clustering analysis.}
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions in the three-dimensional space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). Multiple stable states can exist if feedbacks between development drivers are present. Statistically, this translates into multi-modal distributions and so-called "regimes". The prosperity-environment plane (Fig \ref{fig:cube_ensamble}A), and inequality-environment plane (Fig \ref{fig:cube_ensamble}C) have two pronounced clusters of countries (the movement of countries during the study period stayed within a relatively circumscribed space), while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has one major cluster and one looser grouping. Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), which means their trajectory over the study period did not take them far from where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

Combining the three planes, the trajectories of countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A). First, there was a group composed of countries with relatively high prosperity and ecological footprint; their movements were relatively less erratic and had larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). This cluster can be called "Regime A". Next, there was a group of countries, mainly from Latin America, that seemed to be "trapped" in a region of high inequality during the study period. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the income inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). Within the three-dimensional space, this cluster of countries with high inequality and relatively low prosperity and environmental impact can be called "Regime B". A third group of countries showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow).

If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators, we are able to include more countries (N=140). This static analysis also yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the findings from the dynamic analysis. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact. The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "Global South". Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.

In both the dynamic and the static analysis, decreasing inequality and decreasing environmental impact appear to pull countries in opposite directions (Fig \ref{fig:pca}), providing evidence for the difficulty of simultaneously achieving both goals. In Fig \ref{fig:pca}A, the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences. While per-capita ecological footprint and GNI point in similar directions (countries tended to increase both at the same time), the Gini for income points in the opposite direction. This suggests that countries that reduce inequality do so after increasing in environmental impact and prosperity. In the ordination performed in the larger static sample (Fig \ref{fig:pca}B), we find a similar pattern. In fact, the trajectories analysis found that countries that reduced inequality subsequently increased in environmental impact (Fig \ref{fig:cube_ensamble}A). Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B), although some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact (Fig \ref{fig:cube_ensamble}B, C).

```{=tex}
\begin{figure*}[htp]
\centering
\includegraphics[width = 6.5in, height = 5in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. Three clusters are identified when looking at the historical trajectory of countries (N=59) from 1995-2017 (A). Two clusters are identified when aggregating over time with means and standard deviations regardless of data completeness. This complements the dynamic analysis with one summary statistic over time for 140 countries (B). For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other. Mean values of per-capita ecological footprint (C), Gini for income (D), and per-capita gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

@@

Understanding the patterns and trends in the relationships between prosperity, inequality, and environmental impact is an important topic for sustainability science and development research. In the existing literatures of economic, social, and environmental science, various arguments have been advanced about the relationships between the three goals. Most commonly, mechanisms have been hypothesized for pair-wise relationships, with Kuznets Curve-type relationships being the most widely analyzed. The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the relationship is reversed. The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020][@Pikett:2014] and the environment [@stern2004rise; @stern1996economic ]. For the latter, conclusions have often varied based on the indicator chosen to proxy environmental impact [@MALER201325]. Our analysis did not detect KC-type relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details). Instead, the findings strongly suggest that growing prosperity is almost always associated with worsening inequality and environmental impact.

What is clear from earlier empirical studies of pair-wise relationships of prosperity, inequality, and the environment is that whatever tradeoffs and synergies that could be identified were mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019]. For instance, in his survey of relevant studies, Chancel [-@Chancel:2020] concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics. It is only by considering the economic, social, and environmental dimensions together, and the different ways in which countries have evolved with respect to them, can we arrive at a more comprehensive understanding. In particular, while the prosperity-inequality and prosperity-environment relationships have been widely analyzed (with mixed conclusions), the inequality-environment dynamic has been less explored – and should also be considered with respect to the potentially mediating effects of the third dimension of prosperity. In our findings, this necessity is underscored by the existence of international clusters that moved in distinctively different directions.
